\documentclass[11pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{float}
\usepackage{hyperref}
\usepackage[numbers]{natbib}
\usepackage{enumitem}
\usepackage{caption}

\hypersetup{
  colorlinks=true,
  linkcolor=black,
  citecolor=blue,
  urlcolor=blue
}

\title{\textbf{Report on ``Spatial Mapping with Gaussian Processes and Nonstationary Fourier Features''}}
\author{
  Siu Pan Lau \\
  Zhichao Zhou
}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
This report summarizes the paper ``Spatial mapping with Gaussian processes and nonstationary Fourier features'' by Ton et al.~\citep{ton2018spatial}. The paper explores how to make the Gaussian process model for spatial mapping more flexible by replacing fixed stationary kernels with nonstationary kernels learned through Fourier feature representations. First, we review the relevant background knowledge about Gaussian random fields, stationary covariance functions, and spectral representations. Then, we describe the main methodological approach of this paper, including nonstationary extension, data-driven frequency learning, and regularization strategies. Finally, we summarize the experimental results on Google stock prices and East African land surface temperature, and discuss the advantages, limitations, and possible extensions of the proposed method.
\end{abstract}

\section{Introduction}

Spatial mapping is concerned with learning and predicting quantities that vary
over space or time from a limited number of observations. Examples include
environmental variables, disease risk, temperature fields, and financial time
series. Gaussian process models are a natural tool for this type of problem
because they define a distribution over functions and use a covariance kernel to
describe how observations at different inputs are related. As Ton et al. note,
``The use of covariance kernels is ubiquitous in the field of spatial
statistics'' \citep[p.~59]{ton2018spatial}.

A usual starting point for GP modelling is to choose a stationary kernel, such as
a Mat\'ern or squared exponential kernel. However, this also means that the shape
of the covariance function is mostly fixed before seeing the data. In real spatial
or temporal data, the dependence may not have the same pattern everywhere. The
paper studies this problem by learning nonstationary kernels from data.

The selected paper aims to address this limitation by constructing more flexible
Gaussian process models through nonstationary Fourier feature representations.
The main idea is to learn kernel structure directly from data while retaining the
computational advantages of finite feature representations. This report first
reviews the background on Gaussian random fields, stationary kernels, and
spectral representations. It then summarizes the proposed method, including the
nonstationary extension, data-driven frequency learning, and regularisation.
Finally, it discusses the experiments on Google stock prices and East African
land surface temperature, followed by the strengths, limitations, and possible
extensions of the approach.

\section{Background}

% Add necessary background here.

\subsection{Gaussian Processes}

A random field is a collection of random variables
\begin{equation}
  X = \{X(\mathbf{s}) : \mathbf{s}\in S\},
\end{equation}
indexed by a set $S$ of locations. It is described through its finite-dimensional distributions, that is, through the distributions of vectors of the form
\begin{equation}
  \left(X(\mathbf{s}_1), \ldots, X(\mathbf{s}_d)\right),
  \qquad \mathbf{s}_1,\ldots,\mathbf{s}_d\in S.
\end{equation}
A Gaussian random field is a random field such that every such finite-dimensional random vector has a multivariate Gaussian distribution. Any Gaussian random field is completely determined by its mean function
\begin{equation}
  m(\mathbf{s}) = \mathbb{E}[X(\mathbf{s})]
\end{equation}
and covariance function
\begin{equation}
  C(\mathbf{s},\mathbf{t}) = \operatorname{Cov}(X(\mathbf{s}),X(\mathbf{t})).
\end{equation}
In this report, Gaussian processes are used in this same finite-dimensional sense. Since the covariance function controls spatial dependence, kernel design is central to Gaussian process models.

\subsection{Stationary and Nonstationary Kernels}

A random field is called weakly stationary if its mean is constant and its covariance depends only on the displacement between two locations:
\begin{equation}
  C(\mathbf{s}, \mathbf{t}) = C_0(\mathbf{t}-\mathbf{s}),
\end{equation}
for some function $C_0$. In the weakly stationary case, one often writes
\begin{equation}
  C(\mathbf{h}) = C_0(\mathbf{h}) = C(\mathbf{s}, \mathbf{s}+\mathbf{h}),
\end{equation}
where $\mathbf{h}$ is the spatial lag. If the covariance depends only on the Euclidean distance $\|\mathbf{h}\|$, the model is also isotropic.

The selected paper points out that spatial modelling has often relied on ``a limited class of stationary kernels such as the Mat\'ern or squared exponential,'' which can limit the expressiveness of the model \citep[p.~59]{ton2018spatial}. In contrast, nonstationary kernels allow the covariance structure to vary with location. The paper therefore considers more general spectral representations and proposes a framework to learn complex nonstationary kernel functions directly from data \citep{ton2018spatial}.

\subsection{Fourier Features}

Bochner's theorem gives the spectral view of stationary covariance modelling. In the course material, its meaning is summarized as follows: every continuous stationary covariance function has a spectral representation; conversely, every such spectral representation defines a valid stationary covariance function. Thus, stationary covariance modelling can be done either in the spatial domain, through covariance decay in space, or in frequency space, through the spectral measure or density.

The selected paper connects this spectral view to Fourier features. It states that recent machine learning research has focused on ``spectral representations to model arbitrary stationary kernels'' and on more general representations including nonstationary kernels \citep[p.~59]{ton2018spatial}. The paper then uses ``the connections between Fourier feature representations, Gaussian processes and neural networks'' to develop a framework that can learn complex nonstationary kernel functions directly from data \citep[p.~59]{ton2018spatial}.

\section{Method}

% Summarize the method proposed in the paper.
\subsection{Nonstationary Extension via Yaglom's Theorem}
First, using the Random Fourier Features, write the kernel as an expectation.
Approximate the expectation with Monte Carlo to sample $m$ frequencies. This
gives a feature map. From the feature map, the cost drops from $O(n^3)$ to
$O(nm^2)$ because we only need to invert a $2m \times 2m$ matrix instead of
the $n \times n$ covariance matrix. However, there are two limitations: it
can only be applied to stationary kernels, and the kernel family must be
chosen first.

Therefore, by extending to Yaglom's theorem, a nonstationary kernel is an
integral over pairs of frequencies. Then by applying symmetrising and Monte
Carlo integration, we get a new feature map again. Nonstationarity arises
because $\boldsymbol{\omega}^1 \neq \boldsymbol{\omega}^2$ makes the kernel
depend on $\mathbf{x}_1$ and $\mathbf{x}_2$ separately.
\begin{equation}
  k(\mathbf{x}_1, \mathbf{x}_2) = \int_{\mathbb{R}^D \times \mathbb{R}^D} e^{i(\boldsymbol{\omega}_1^\top \mathbf{x}_1 - \boldsymbol{\omega}_2^\top \mathbf{x}_2)} \, \mu(\mathrm{d}\boldsymbol{\omega}_1, \mathrm{d}\boldsymbol{\omega}_2).
\end{equation}

\begin{equation}
  \Phi_x = \left[ \cos(X \Omega_1^\top) + \cos(X \Omega_2^\top) \;\big|\; \sin(X \Omega_1^\top) + \sin(X \Omega_2^\top) \right] \in \mathbb{R}^{n \times 2m}.
\end{equation}
Vanilla RFF samples frequencies from a fixed distribution. It would force us
to choose a kernel family. Instead, we can treat the frequency pairs as free
parameters and learn them from data. We can optimise the frequencies by ADAM
gradient descent on the log marginal likelihood. Each iteration uses the
current frequencies to build the feature map and compute the log marginal
likelihood. Then backpropagate to get the gradient and update the frequencies
again. $\Omega$ plays the role of the weights. However, with many parameters
and a non-convex objective, the model can overfit.
\subsection{Learning Frequencies from Data}
\begin{equation}
  \log p(\mathbf{y}|\theta) = -\frac{1}{2\sigma_n^2}\left(\|\mathbf{y}\|^2 - \|\alpha_1\|^2\right) - \frac{1}{2}\sum_i \log R_{ii}^2 + m\log\frac{m\sigma_n^2}{\sigma_f^2} - \frac{n}{2}\log(2\pi \sigma_n^2),
\end{equation}
where $R = \operatorname{chol}(A)$, $A = \Phi_x^\top \Phi_x + \tfrac{m\sigma_n^2}{\sigma_f^2}\, I_{2m}$, and $\alpha_1$ solves $R\alpha_1 = \Phi_x^\top \mathbf{y}$.

\begin{equation}
  \Omega \leftarrow \Omega + \eta \cdot \mathrm{ADAM}\!\left( \frac{\partial \log p(\mathbf{y}|\theta)}{\partial \Omega} \right),
\end{equation}
where $\eta$ is the learning rate and the gradient is computed automatically via backpropagation through the feature map.

\subsection{Regularisation via Gaussian Dropout}

Standard dropout does not work well here because of the cos/sin
transformations. Gaussian dropout \citep{srivastava2014dropout,
baldi2013understanding} is used instead: multiply $\Omega$ by element-wise
noise drawn from $\mathcal{N}(1, \sigma_p^2)$. This preserves the expected
value of the covariance matrix, prevents co-adaptation of frequencies, and
improves robustness to noise. As additional regularisation, early stopping
\citep{prechelt1998early} monitors the cross-validation error. The dropout
noise $\sigma_p$ and the learning rate are chosen by random search over 500
samples.
\begin{equation}
  \{(\boldsymbol{\omega}^1_k, \boldsymbol{\omega}^2_k)_\eta\}_{k=1}^{m} = \mathcal{N}(1, \sigma_p^2) \odot \{(\boldsymbol{\omega}^1_k, \boldsymbol{\omega}^2_k)\}_{k=1}^{m},
\end{equation}
where $\odot$ denotes elementwise multiplication and $\sigma_p$ controls the regularisation strength.


\section{Experiments}

% Summarize the experiments from the paper.

\begin{table}[H]
  \centering
  \caption{Summary of experiments reported in Ton et al.~\citep[sec.~4]{ton2018spatial}.}
  \begin{tabular}{p{0.28\linewidth}p{0.33\linewidth}p{0.30\linewidth}}
    \toprule
    Task & Data & Main result \\
    \midrule
    Google stock price &
    3295 daily observations from 19 August 2004 to 20 September 2017. &
    Proposed method: MSE $3.29\times10^{-4}$, correlation $0.999$. \\
    East Africa temperature anomaly &
    MOD11A2 land surface temperature data with 6000 sampled locations. &
    Proposed method and MRA: MAE $3.5$; SPDE: $3.9$; LatticeKrig: $4.3$. \\
    \bottomrule
  \end{tabular}
\end{table}

\section{Discussion}

% Discuss strengths, limitations, and your group's interpretation.

\subsection{Strengths}

\subsubsection{No additional computational cost}
Based on the method, the cost drops from $O(n^3)$ to $O(nm^2)$ with $m \ll n$.
It helps to reduce the computation time. Real-world spatial datasets can have
millions of points, and this improves the time efficiency, while also lowering
computer memory requirements.

\subsubsection{Kernel learned directly from data}
% Add strengths here.
In the traditional approach, the hyperparameters need to be guessed before
processing. For example,
\begin{equation}
  k(\mathbf{x}_1, \mathbf{x}_2) = \sigma_f^2 \exp\!\left(-\frac{\|\mathbf{x}_1 - \mathbf{x}_2\|^2}{2\ell^2}\right).
\end{equation}
With this paper's approach, no kernel family needs to be chosen anymore. The
frequencies $\boldsymbol{\omega}^1_k, \boldsymbol{\omega}^2_k$ are learned
from the data directly.

\subsubsection{Nonstationarity}
Traditional Gaussian process regression requires many assumptions. It can
only be used for stationary kernels. This paper's proposed method can also be
used for nonstationary kernels, which more precisely simulate and predict
real-life applications, just like the temperature in East Africa shown in
the experiment.
\subsection{Limitations}

% Add limitations here.
\subsubsection{Fine-scale spatial detail not fully captured}
\subsection{Possible Extensions}

% Add possible extensions here.

\section{Conclusion}

% Write the conclusion here.

\bibliographystyle{unsrtnat}
\bibliography{references}

\end{document}
